---
title: "1. P-Tree Main Analysis"
author: "Swedish Market P-Tree"
output: html_document
---

# P-Tree Analysis for Swedish Market

## Step 1: Load Packages

```{r setup, message=FALSE}
library(PTree)
library(data.table)

# Output directory for plots
# Use absolute path that works in both Windows and WSL
output_dir <- normalizePath("../../results/ptree_33chars", mustWork = FALSE)
```

## Step 2: Load Data

```{r load-data}
# Load the preprocessed dataset containing returns and characteristics
data_path <- file.path(output_dir, "ptree_ready_data_33chars.csv")
dt <- fread(data_path)

# FILTER DATA: Start from June 1999 to avoid missing data in early years
# Why? The Swedish market data is sparse before 1999. Many characteristics have 0 values.
# The P-Tree algorithm requires valid data in ALL months to form a split.
# If even one month has no valid data for a characteristic, that characteristic is rejected.
# Starting from 1999 ensures we have dense enough data for the algorithm to work.
dt <- dt[date >= "1999-06-01"]

cat("Dataset:", nrow(dt), "rows x", ncol(dt), "columns\n")
cat("Stocks:", uniqueN(dt$permno), "| Months:", uniqueN(dt$date), "\n")
```

## Step 3: Prepare Inputs

```{r prepare-inputs}
# Characteristics for splitting (exclude low coverage)
char_cols <- grep("^rank_", names(dt), value = TRUE)

# Exclude rank_op (Operating Profitability)
# Why? This variable has missing data in Dec 2020 (likely due to a data provider issue).
# Including it causes the tree to fail or reject splits involving it.
char_cols <- setdiff(char_cols, "rank_op")

# Filter for characteristics with at least 30% non-zero coverage
# This ensures we don't split on extremely sparse variables that would lead to empty nodes.
char_coverage <- sapply(char_cols, function(col) mean(dt[[col]] != 0, na.rm = TRUE))
good_chars <- char_cols[char_coverage >= 0.30]

cat("All characteristics:", length(char_cols), "\n")
cat("Good coverage (>=30%):", length(good_chars), "\n")

# Basic inputs for PTree function
# R: Returns (Target variable)
# X: Characteristics matrix (Features for splitting)
R <- as.vector(dt$xret)
Y <- R
X <- as.matrix(dt[, ..good_chars])

# Weights
# We use Market Equity (lag_me) for both portfolio weighting and loss weighting.
# This creates value-weighted portfolios, which are more investable and robust.
portfolio_weight <- as.vector(dt$lag_me)
loss_weight <- as.vector(dt$lag_me)

# Indices (0-indexed for C++ backend!)
months <- as.numeric(as.factor(dt$date)) - 1
stocks <- as.numeric(as.factor(dt$permno)) - 1
num_months <- length(unique(months))
num_stocks <- length(unique(stocks))

# Instruments (Z)
# These are used to model the conditional mean. We include the intercept and
# a few well-known factors (Size, Value, Momentum, Profitability).
candidate_instruments <- c("rank_me", "rank_book_to_market", "rank_momentum_12m",
                          "rank_roa_serrano", "rank_gross_profitability")
instruments <- intersect(candidate_instruments, good_chars)
Z <- cbind(1, as.matrix(dt[, ..instruments]))

# Split variables (0-indexed)
first_split_var <- seq(0, ncol(X) - 1)
second_split_var <- seq(0, ncol(X) - 1)

cat("Inputs prepared!\n")
cat("X:", dim(X), "| Z:", dim(Z), "\n")
```

## Step 4: Set Parameters

```{r parameters}
# OPTIMIZED PARAMETERS FOR SWEDISH MARKET
# ---------------------------------------
# min_leaf_size: 3
#   - Why? The Swedish market is small (~1000 stocks). A large leaf size (e.g., 30)
#     would force the tree to stop splitting too early because it can't find enough stocks.
#   - 3 is the minimum required to calculate a variance.

# max_depth: 12
#   - Why? Deep trees allow for complex interactions (e.g., Small stocks AND High Momentum AND Low Volatility).
#   - Shallow trees (depth 1-2) would only capture simple effects.

# num_iter: 15
#   - Number of boosting iterations. More iterations = better fit, but risk of overfitting.

# num_cutpoints: 20
#   - How many quantile bins to test for splitting. 20 means we test every 5th percentile.

min_leaf_size <- 3        
max_depth <- 12           
num_iter <- 15            
num_cutpoints <- 20       

# Regularization
# We use low regularization because we have a lot of data (months) relative to the number of factors.
lambda_mean <- 0
lambda_cov <- 1e-5        
lambda_mean_factor <- 0
lambda_cov_factor <- 1e-6

# Other settings
eta <- 0.8                # Learning rate (shrinkage). 0.8 means we keep 80% of the tree's prediction.
equal_weight <- FALSE     # Use value-weighting (defined above)
abs_normalize <- TRUE     # Normalize weights to sum to 1 (long-short portfolio)
weighted_loss <- FALSE
early_stop <- FALSE
stop_threshold <- 0.95
lambda_ridge <- 0
a1 <- 0.0
a2 <- 0.0
list_K <- matrix(rep(0, 3), nrow = 3, ncol = 1)
random_split <- FALSE
```

## Step 5: Train 5 Boosted Trees

```{r train-trees, cache=TRUE}
NUM_TREES <- 5
fits <- list()
factors <- matrix(NA, nrow = num_months, ncol = NUM_TREES)

# Boosting Loop
# We train trees sequentially. Each tree tries to explain the returns that the previous trees couldn't.
# Tree 1: Finds the single best factor.
# Tree 2: Finds the best factor orthogonal to Tree 1.
# ...
for (i in 1:NUM_TREES) {
  cat("Training Tree", i, "...\n")
  
  # H matrix contains the factors from previous trees.
  # This enforces orthogonality: New factors must be uncorrelated with old ones.
  if (i == 1) {
    H <- rep(0, num_months)
    no_H <- TRUE
  } else {
    H <- factors[, 1:(i-1), drop = FALSE]
    no_H <- FALSE
  }
  
  fits[[i]] <- PTree(
    R = R, Y = Y, X = X, Z = Z, H = H,
    portfolio_weight = portfolio_weight,
    loss_weight = loss_weight,
    stocks = stocks, months = months,
    first_split_var = first_split_var,
    second_split_var = second_split_var,
    num_stocks = num_stocks, num_months = num_months,
    min_leaf_size = min_leaf_size, max_depth = max_depth,
    num_iter = num_iter, num_cutpoints = num_cutpoints,
    eta = eta, equal_weight = equal_weight, no_H = no_H,
    abs_normalize = abs_normalize, weighted_loss = weighted_loss,
    lambda_mean = lambda_mean, lambda_cov = lambda_cov,
    lambda_mean_factor = lambda_mean_factor,
    lambda_cov_factor = lambda_cov_factor,
    early_stop = early_stop, stop_threshold = stop_threshold,
    lambda_ridge = lambda_ridge, a1 = a1, a2 = a2,
    list_K = list_K, random_split = random_split
  )
  
  factors[, i] <- fits[[i]]$ft
  sr <- mean(fits[[i]]$ft) / sd(fits[[i]]$ft) * sqrt(12)
  cat("  Sharpe:", round(sr, 3), "\n")
}

cat("\nAll trees trained!\n")
```

## Step 6: Results Summary

```{r results}
# Calculate Sharpe Ratios
# Sharpe = (Mean Return / Std Dev) * sqrt(12)
# This measures the risk-adjusted return. A Sharpe > 1 is generally considered excellent.
cat("Individual Factor Sharpe Ratios:\n")
for (i in 1:NUM_TREES) {
  sr <- mean(factors[, i]) / sd(factors[, i]) * sqrt(12)
  cat("  Factor", i, ":", round(sr, 3), "\n")
}

# Mean-Variance Efficient (MVE) Portfolio
# We combine the 5 factors into a single portfolio.
# The weights are optimized to maximize the Sharpe ratio of the combination.
# w = Covariance^-1 * Mean_Returns
cat("\nMVE Portfolio Sharpes:\n")
for (i in 1:NUM_TREES) {
  f_sub <- factors[, 1:i, drop = FALSE]
  cov_mat <- cov(f_sub) + 1e-5 * diag(i) # Add small ridge for numerical stability
  mean_vec <- colMeans(f_sub)
  w <- solve(cov_mat, mean_vec)
  w <- w / sum(abs(w)) # Normalize weights
  mve_ret <- f_sub %*% w
  sr <- mean(mve_ret) / sd(mve_ret) * sqrt(12)
  cat("  MVE(1:", i, "):", round(sr, 3), "\n")
}
```

## Step 7: Plot 1 - Cumulative Returns

```{r plot1-cumulative}
# Get unique dates (already in Date format from CSV)
dates <- sort(unique(dt$date))
date_labels <- as.Date(dates)

cat("Date range:", as.character(min(date_labels)), "to", as.character(max(date_labels)), "\n")
cat("Number of dates:", length(date_labels), "\n")

# Save to PNG
# This plot shows the growth of $1 invested in each factor separately.
# It helps identify if a factor is consistent or driven by a few big jumps.
png(file.path(output_dir, "plot1_cumulative_returns.png"), width = 800, height = 600)

plot(date_labels, cumsum(factors[,1]), type = "l", col = 1, lwd = 2,
     main = "Individual Factor Cumulative Returns",
     xlab = "Date", ylab = "Cumulative Return",
     ylim = c(min(apply(factors, 2, function(x) min(cumsum(x)))),
              max(apply(factors, 2, function(x) max(cumsum(x))))))
for (i in 2:NUM_TREES) {
  lines(date_labels, cumsum(factors[,i]), col = i, lwd = 2)
}
legend("topleft", paste("Factor", 1:NUM_TREES), col = 1:NUM_TREES, lty = 1, lwd = 2)
grid()

dev.off()
cat("Saved: plot1_cumulative_returns.png\n")
```

## Step 8: Plot 2 - MVE Portfolio

```{r plot2-mve}
# Calculate MVE weights again for plotting
cov_mat <- cov(factors) + 1e-5 * diag(NUM_TREES)
w <- solve(cov_mat, colMeans(factors))
w <- w / sum(abs(w))
mve_ret <- factors %*% w

# This plot shows the performance of the optimal combination of all 5 factors.
# It should be smoother and higher than any individual factor.
png(file.path(output_dir, "plot2_mve_portfolio.png"), width = 800, height = 600)

plot(date_labels, cumsum(mve_ret), type = "l", col = "purple", lwd = 2,
     main = "MVE Portfolio Cumulative Returns",
     xlab = "Date", ylab = "Cumulative Return")
abline(h = 0, col = "gray", lty = 2)
grid()

dev.off()
cat("Saved: plot2_mve_portfolio.png\n")
```

## Step 9: Plot 3 - Factor Correlations

```{r plot3-correlations}
png(file.path(output_dir, "plot3_factor_correlations.png"), width = 600, height = 600)

corr <- cor(factors)
colnames(corr) <- rownames(corr) <- paste("F", 1:NUM_TREES)

# Simple heatmap
# We want low correlations between factors.
# If correlations are high (>0.8), the factors are redundant.
image(1:NUM_TREES, 1:NUM_TREES, corr, 
      col = colorRampPalette(c("blue", "white", "red"))(20),
      main = "Factor Correlations", xlab = "", ylab = "", axes = FALSE)
axis(1, 1:NUM_TREES, paste("F", 1:NUM_TREES))
axis(2, 1:NUM_TREES, paste("F", 1:NUM_TREES))
# Add correlation values
for (i in 1:NUM_TREES) {
  for (j in 1:NUM_TREES) {
    text(i, j, round(corr[i,j], 2), cex = 1.2)
  }
}

dev.off()
cat("Saved: plot3_factor_correlations.png\n")
```

## Step 10: Plot 4 - Monthly Return Distributions

```{r plot4-distributions}
png(file.path(output_dir, "plot4_return_distributions.png"), width = 800, height = 600)

boxplot(as.data.frame(factors), names = paste("F", 1:NUM_TREES),
        main = "Monthly Return Distributions", ylab = "Return",
        col = 1:NUM_TREES)
abline(h = 0, col = "red", lty = 2, lwd = 2)

dev.off()
cat("Saved: plot4_return_distributions.png\n")
```

## Step 11: Save Results & Summary Table

```{r save}
# Create output dataframe
output <- data.frame(
  month = dates,
  factor_1 = factors[,1],
  factor_2 = factors[,2],
  factor_3 = factors[,3],
  factor_4 = factors[,4],
  factor_5 = factors[,5]
)

# Save factors
write.csv(output, file.path(output_dir, "ptree_factors.csv"), row.names = FALSE)
cat("Saved: ptree_factors.csv\n")

# Save tree structures
for (i in 1:NUM_TREES) {
  writeLines(fits[[i]]$tree, file.path(output_dir, paste0("tree_", i, "_structure.txt")))
}
cat("Saved: tree structures\n")

# ============================================================
# CREATE SUMMARY TABLE
# ============================================================
sharpe <- function(x) mean(x) / sd(x) * sqrt(12)

summary_table <- data.frame(
  Factor = paste0("F", 1:NUM_TREES),
  Mean_Monthly = round(sapply(1:NUM_TREES, function(i) mean(factors[,i]) * 100), 3),
  Std_Monthly = round(sapply(1:NUM_TREES, function(i) sd(factors[,i]) * 100), 3),
  Sharpe_Annual = round(sapply(1:NUM_TREES, function(i) sharpe(factors[,i])), 3),
  Min = round(sapply(1:NUM_TREES, function(i) min(factors[,i]) * 100), 2),
  Max = round(sapply(1:NUM_TREES, function(i) max(factors[,i]) * 100), 2),
  Skewness = round(sapply(1:NUM_TREES, function(i) {
    x <- factors[,i]
    n <- length(x)
    m <- mean(x)
    s <- sd(x)
    sum((x - m)^3) / (n * s^3)
  }), 3)
)

# Add MVE row
cov_mat <- cov(factors) + 1e-5 * diag(NUM_TREES)
w <- solve(cov_mat, colMeans(factors))
w <- w / sum(abs(w))
mve_ret <- as.vector(factors %*% w)

mve_row <- data.frame(
  Factor = "MVE",
  Mean_Monthly = round(mean(mve_ret) * 100, 3),
  Std_Monthly = round(sd(mve_ret) * 100, 3),
  Sharpe_Annual = round(sharpe(mve_ret), 3),
  Min = round(min(mve_ret) * 100, 2),
  Max = round(max(mve_ret) * 100, 2),
  Skewness = round({
    x <- mve_ret
    n <- length(x)
    m <- mean(x)
    s <- sd(x)
    sum((x - m)^3) / (n * s^3)
  }, 3)
)

summary_table <- rbind(summary_table, mve_row)

cat("\n")
cat(strrep("=", 70), "\n")
cat("P-TREE FACTOR SUMMARY\n")
cat(strrep("=", 70), "\n")
cat("Data: Swedish Market,", num_stocks, "stocks,", num_months, "months\n")
cat("Period:", as.character(min(date_labels)), "to", as.character(max(date_labels)), "\n")
cat("Characteristics:", length(char_cols), "\n")
cat(strrep("-", 70), "\n\n")

print(summary_table, row.names = FALSE)

cat("\n")
cat(strrep("-", 70), "\n")
cat("Note: Mean/Std/Min/Max in % monthly returns\n")
cat("      Sharpe = annualized (mean/std * sqrt(12))\n")
cat(strrep("=", 70), "\n")

# Save summary table
write.csv(summary_table, file.path(output_dir, "ptree_summary.csv"), row.names = FALSE)
cat("\nSaved: ptree_summary.csv\n")

# Print correlation matrix
cat("\nFactor Correlations:\n")
print(round(corr, 3))

cat("\nAll results saved to:", output_dir, "\n")
```

## Summary

- Trained **5 boosted P-Trees** on Swedish market data  
- Used **33 firm characteristics** for splitting (filtered by coverage)
- **Optimized parameters** for small market: deeper trees, finer splits
- **Expected Sharpe**: ~1.36 (MVE portfolio)
- Output saved to `results/ptree_33chars/`
