---
title: "2. Benchmark Analysis"
author: "P-Tree vs Fama-French Factors"
output: html_document
---

# Benchmark Analysis

Compare P-Tree factors against standard benchmarks (CAPM, FF3, FF4).

## Step 1: Load Packages

```{r setup, message=FALSE}
library(data.table)

# Install if needed
if (!require(sandwich)) install.packages("sandwich")
if (!require(lmtest)) install.packages("lmtest")
library(sandwich)
library(lmtest)

output_dir <- "c:/Users/micha/Desktop/Projekte/PTrees/results/ptree_33chars"
```

## Step 2: Load Data

```{r load-data}
# Load P-Tree factors (from Step 1)
ptree <- fread(file.path(output_dir, "ptree_factors.csv"))
ptree$date <- as.Date(ptree$month)  # month column is already a date
# Create year-month for merging
ptree$ym <- format(ptree$date, "%Y-%m")
cat("P-Tree factors:", nrow(ptree), "months\n")

# Load macro/benchmark factors (Swedish Fama-French)
macro <- fread("c:/Users/micha/Desktop/Projekte/PTrees/data/raw/FamaFrench2020/FF4F_monthly.csv")
cat("Macro data:", nrow(macro), "months\n")

# Merge on year-month
merged <- merge(ptree, macro, by = "ym")
cat("Overlapping:", nrow(merged), "months\n")
cat("Period:", min(merged$ym), "to", max(merged$ym), "\n")
```

## Step 3: Extract Factors

```{r extract-factors}
# P-Tree factors
f1 <- merged$factor_1
f2 <- merged$factor_2
f3 <- merged$factor_3
f4 <- merged$factor_4
f5 <- merged$factor_5

# Benchmark factors
mkt <- merged$rm_rf
smb <- merged$smb_vw
hml <- merged$hml_vw
mom <- merged$mom_vw

cat("Factors extracted\n")
```

## Step 4: Sharpe Ratios

```{r sharpe}
sharpe <- function(x) mean(x, na.rm = TRUE) / sd(x, na.rm = TRUE) * sqrt(12)

cat(strrep("=", 60), "\n")
cat("SHARPE RATIOS (Annualized)\n")
cat(strrep("=", 60), "\n\n")

cat("P-Tree Factors:\n")
cat("  F1:", round(sharpe(f1), 3), "\n")
cat("  F2:", round(sharpe(f2), 3), "\n")
cat("  F3:", round(sharpe(f3), 3), "\n")
cat("  F4:", round(sharpe(f4), 3), "\n")
cat("  F5:", round(sharpe(f5), 3), "\n")

cat("\nBenchmark Factors:\n")
cat("  MKT:", round(sharpe(mkt), 3), "\n")
cat("  SMB:", round(sharpe(smb), 3), "\n")
cat("  HML:", round(sharpe(hml), 3), "\n")
cat("  MOM:", round(sharpe(mom), 3), "\n")

# MVE combination
factors_mat <- cbind(f1, f2, f3, f4, f5)
cov_mat <- cov(factors_mat) + 1e-5 * diag(5)
w <- solve(cov_mat, colMeans(factors_mat))
w <- w / sum(abs(w))
mve_ret <- factors_mat %*% w

cat("\nMVE Portfolio (all 5 P-Tree):\n")
cat("  Sharpe:", round(sharpe(mve_ret), 3), "\n")
```

## Step 5: Alpha Regressions

```{r alpha-regressions}
run_alpha_reg <- function(y, X, name) {
  model <- lm(y ~ X)
  hac_se <- coeftest(model, vcov = vcovHAC(model, lag = 3))
  
  alpha <- coef(model)[1] * 12 * 100
  t_stat <- hac_se[1, 3]
  r2 <- summary(model)$r.squared
  
  cat(sprintf("  %s: alpha = %6.2f%% (t = %5.2f), R2 = %.3f\n", name, alpha, t_stat, r2))
  return(c(alpha = alpha, t = t_stat, r2 = r2))
}

cat("\n", strrep("=", 60), "\n")
cat("ALPHA VS BENCHMARKS (Annualized %, HAC t-stats)\n")
cat(strrep("=", 60), "\n")

for (i in 1:5) {
  f <- get(paste0("f", i))
  cat("\nFactor", i, ":\n")
  run_alpha_reg(f, mkt, "CAPM")
  run_alpha_reg(f, cbind(mkt, smb, hml), "FF3 ")
  run_alpha_reg(f, cbind(mkt, smb, hml, mom), "FF4 ")
}
```

## Step 6: Correlations

```{r correlations}
cat("\n", strrep("=", 60), "\n")
cat("FACTOR CORRELATIONS\n")
cat(strrep("=", 60), "\n\n")

all_factors <- data.frame(
  F1 = f1, F2 = f2, F3 = f3, F4 = f4, F5 = f5,
  MKT = mkt, SMB = smb, HML = hml, MOM = mom
)

corr <- round(cor(all_factors, use = "complete.obs"), 3)
print(corr)
```

## Step 7: Summary Table

```{r summary-table}
cat("\n", strrep("=", 60), "\n")
cat("SUMMARY TABLE\n")
cat(strrep("=", 60), "\n\n")

results <- data.frame(
  Factor = paste0("F", 1:5),
  Sharpe = NA, Alpha_CAPM = NA, t_CAPM = NA, Alpha_FF4 = NA, t_FF4 = NA
)

for (i in 1:5) {
  f <- get(paste0("f", i))
  results$Sharpe[i] <- round(sharpe(f), 3)
  
  m1 <- lm(f ~ mkt)
  h1 <- coeftest(m1, vcov = vcovHAC(m1, lag = 3))
  results$Alpha_CAPM[i] <- round(coef(m1)[1] * 12 * 100, 2)
  results$t_CAPM[i] <- round(h1[1, 3], 2)
  
  m4 <- lm(f ~ mkt + smb + hml + mom)
  h4 <- coeftest(m4, vcov = vcovHAC(m4, lag = 3))
  results$Alpha_FF4[i] <- round(coef(m4)[1] * 12 * 100, 2)
  results$t_FF4[i] <- round(h4[1, 3], 2)
}

print(results)

write.csv(results, file.path(output_dir, "benchmark_results.csv"), row.names = FALSE)
cat("\nSaved: benchmark_results.csv\n")
```

## Step 8: Plot 5 - P-Tree vs Market

```{r plot5-vs-market}
# Create date sequence for plotting
plot_dates <- as.Date(paste0(merged$ym, "-01"))

png(file.path(output_dir, "plot5_ptree_vs_market.png"), width = 800, height = 600)

plot(plot_dates, cumsum(f1), type = "l", col = "blue", lwd = 2,
     main = "P-Tree Factor 1 vs Market",
     xlab = "Date", ylab = "Cumulative Return")
lines(plot_dates, cumsum(mkt), col = "black", lty = 2, lwd = 2)
legend("topleft", c("P-Tree F1", "Market"), col = c("blue", "black"), lty = c(1, 2), lwd = 2)
grid()

dev.off()
cat("Saved: plot5_ptree_vs_market.png\n")
```

## Step 9: Plot 6 - MVE vs Market

```{r plot6-mve-vs-market}
png(file.path(output_dir, "plot6_mve_vs_market.png"), width = 800, height = 600)

plot(plot_dates, cumsum(mve_ret), type = "l", col = "purple", lwd = 2,
     main = "P-Tree MVE vs Market",
     xlab = "Date", ylab = "Cumulative Return")
lines(plot_dates, cumsum(mkt), col = "black", lty = 2, lwd = 2)
legend("topleft", c("P-Tree MVE", "Market"), col = c("purple", "black"), lty = c(1, 2), lwd = 2)
grid()

dev.off()
cat("Saved: plot6_mve_vs_market.png\n")
```

## Step 10: Plot 7 - Full Correlation Heatmap

```{r plot7-full-correlations}
png(file.path(output_dir, "plot7_full_correlations.png"), width = 700, height = 700)

n <- ncol(all_factors)
image(1:n, 1:n, as.matrix(corr), 
      col = colorRampPalette(c("blue", "white", "red"))(20),
      main = "P-Tree vs Benchmark Correlations", 
      xlab = "", ylab = "", axes = FALSE,
      zlim = c(-1, 1))
axis(1, 1:n, colnames(corr), las = 2, cex.axis = 0.8)
axis(2, 1:n, colnames(corr), las = 2, cex.axis = 0.8)
for (i in 1:n) {
  for (j in 1:n) {
    text(i, j, corr[i,j], cex = 0.8)
  }
}

dev.off()
cat("Saved: plot7_full_correlations.png\n")
```

## Step 11: Plot 8 - Sharpe Comparison

```{r plot8-sharpe-comparison}
png(file.path(output_dir, "plot8_sharpe_comparison.png"), width = 800, height = 600)

sharpes <- c(
  sapply(1:5, function(i) sharpe(get(paste0("f", i)))),
  sharpe(mkt), sharpe(smb), sharpe(hml), sharpe(mom)
)
names(sharpes) <- c(paste0("F", 1:5), "MKT", "SMB", "HML", "MOM")

barplot(sharpes, main = "Sharpe Ratios: P-Tree vs Benchmarks", 
        col = c(rep("steelblue", 5), rep("gray", 4)),
        las = 2, ylab = "Annualized Sharpe Ratio")
abline(h = 0)

dev.off()
cat("Saved: plot8_sharpe_comparison.png\n")
```

## Summary

**Key Findings:**

1. **Sharpe Ratios**: P-Tree factors vs benchmarks
2. **Alpha**: Excess returns after controlling for known factors  
3. **Correlations**: How factors relate to each other

All results and plots saved to `results/ptree_33chars/`
